---
title: "Plot predictor results"
author: "Shraddha Pai"
package: netDx
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
    %\VignetteEngine{knitr::knitr}
    %\VignetteIndexEntry{02. Plot predictor results}
    %\VignetteEncoding{UTF-8}
---

# TL;DR
```{r,eval=FALSE}
suppressMessages(library(netDx))
suppressMessages(library(curatedTCGAData))
suppressMessages(library(MultiAssayExperiment))
suppressMessages(library(TCGAutils))

# load example predictor output data
inDir <- sprintf("%s/extdata/example_output",
	path.package("netDx"))
all_rngs <- list.dirs(inDir, recursive = FALSE)

predClasses <- c("LumA","notLumA")

# plot model performance
predFiles <- unlist(lapply(all_rngs, function(x) 
		paste(x, "predictionResults.txt", sep = "/")))
predPerf <- plotPerf(inDir, predClasses=predClasses)

# list top-scoring features 
featScores <- getFeatureScores(inDir,predClasses=c("LumA","notLumA"))
featSelNet <- lapply(featScores, function(x) {
	callFeatSel(x, fsCutoff=2, fsPctPass=0.7)
})

# plot network view of top-scoring pathways (EnrichmentMap)
# requires Cytoscape to be open - commented out here
###
### # data setup  - fetch data to limit pathways to genes assayed in the data
### brca <- suppressMessages(curatedTCGAData("BRCA",c("mRNAArray"),FALSE))
### xpr_genes <- rownames(assays(brca)[[1]])
### 
### #setup
###pathwayList <- readPathways(getExamplePathways())
###pathwayList <- lapply(pathwayList, function(x) x[which(x %in% xpr_genes)])
###
###netInfoFile <- sprintf("%s/inputNets.txt",inDir)
###netInfo <- read.delim(netInfoFile,sep="\t",h=TRUE,as.is=TRUE)
###
#### fetch input to create EnrichmentMap
###Emap_res <- getEMapInput_many(featScores,pathwayList,
###	maxScore=10,pctPass=0.7,netInfo,verbose=FALSE)
###
# write emap results to file 
###gmtFiles <- list()
###nodeAttrFiles <- list()
###
###for (g in names(Emap_res)) {
###	outFile <- sprintf("%s/%s_nodeAttrs.txt",outDir,g)
###	write.table(Emap_res[[g]][["nodeAttrs"]],file=outFile,
###		sep="\t",col=TRUE,row=FALSE,quote=FALSE)
###	nodeAttrFiles[[g]] <- outFile
###
###	outFile <- sprintf("%s/%s.gmt",outDir,g)
###	conn <- base::file(outFile,"w")
###	tmp <- Emap_res[[g]][["featureSets"]]
###	gmtFiles[[g]] <- outFile
###
###	for (cur in names(tmp)) {
###		curr <- sprintf("%s\t%s\t%s", cur,cur,
###			paste(tmp[[cur]],collapse="\t"))
###		writeLines(curr,con=conn)
###	}
###close(conn)
###}
#### plot EnrichmentMap - needs Cytoscape to be open
###plotEmap(gmtFiles[[1]],nodeAttrFiles[[1]])
```

# Setup

```{r}
suppressMessages(require(netDx))

suppressMessages(library(curatedTCGAData))
suppressMessages(library(MultiAssayExperiment))
suppressMessages(library(TCGAutils))
```

# Prepare data 

Load predictor results generated by running `buildPredictor()`. Here we have
saved the results so the example runs quickly and simple load these results.

```{r}
# load example predictor output data
inDir <- sprintf("%s/extdata/example_output",
	path.package("netDx"))
all_rngs <- list.dirs(inDir, recursive = FALSE)

predClasses <- c("LumA","notLumA")
```

Plot distribution of AUROC and AUPR, as well as ROC and precision-recall curves showing performance across splits.

```{r}
# plot model performance
predFiles <- unlist(lapply(all_rngs, function(x) 
		paste(x, "predictionResults.txt", sep = "/")))
predPerf <- plotPerf(inDir, predClasses=predClasses)
```

Get feature scores for all train/test splits
```{r}
featScores <- getFeatureScores(inDir,predClasses=c("LumA","notLumA"))

print(names(featScores))
```

Look at scores for the LumA label:

```{r}
head(featScores$LumA[,1:3])
```

NA values indicate a score of zero.

Now get the features that scored at least 2 out of 2 in over 70% of the splits.

```{r}
featSelNet <- lapply(featScores, function(x) {
	callFeatSel(x, fsCutoff=2, fsPctPass=0.7)
})
print(head(featSelNet$LumA))
```


# sessionInfo

```{r}
sessionInfo()
```
