---
title: "Validate trained model on an independent dataset"
author: "Shraddha Pai"
package: netDx
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
    %\VignetteIndexEntry{04. Validate model with selected features on an independent dataset}.
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
knitr::opts_chunk$set(crop=NULL)
```

# Introduction

Validating a trained model on an independent new dataset is important to ensure that the model generalizes on new samples and has real-world value. In netDx, models are trained using `buildPredictor()`, which also scores features based on their predictive value for various patient labels. Therafter, we use the `predict()` function to classify samples held out from the original training. You can just as easily use samples from an independent dataset. 

As with the earlier vignette, we will use the application of classifying breast tumour profiles as either "Luminal A" or "other" subtype, by combining clinical and transcriptomic data. 

# Split Data into Training and Holdout Set

In this example we will separate out 20 samples of the data (10 per class) to validate our trained model; those samples will not be used to train the model. We start by fetching the multi-omic dataset using the `curatedTCGAData` package, performing some data formatting, and then finally separating the holdout set (here, `holdout`). 

```{r,eval=TRUE}
suppressWarnings(suppressMessages(require(netDx)))
suppressMessages(require(curatedTCGAData))

brca <- suppressMessages(
  curatedTCGAData("BRCA",c("mRNAArray"),FALSE, version="1.1.38"))

staget <- sub("[abcd]","",
  sub("t","",colData(brca)$pathology_T_stage))
staget <- suppressWarnings(as.integer(staget))
colData(brca)$STAGE <- staget

pam50 <- colData(brca)$PAM50.mRNA
pam50[which(!pam50 %in% "Luminal A")] <- "notLumA"
pam50[which(pam50 %in% "Luminal A")] <- "LumA"
colData(brca)$pam_mod <- pam50

tmp <- colData(brca)$PAM50.mRNA
idx <- union(which(tmp %in% c("Normal-like",
                  "Luminal B","HER2-enriched")),
            which(is.na(staget)))
pID <- colData(brca)$patientID
tokeep <- setdiff(pID, pID[idx])
brca <- brca[,tokeep,]

# remove duplicate assays mapped to the same sample
smp <- sampleMap(brca)
samps <- smp[which(smp$assay=="BRCA_mRNAArray-20160128"),]
notdup <- samps[which(!duplicated(samps$primary)),"colname"]
brca[[1]] <- suppressMessages(brca[[1]][,notdup])

pam50 <- colData(brca)$pam_mod
pheno <- colData(brca)
```

We next set 20 samples aside for independent validation. In practice, this sample size may be larger (e.g. 30% of the samples are held out), or an independent dataset may be used for the same.

```{r,eval=TRUE}
set.seed(123) # make reproducible
idx_holdout <- c(
  sample(which(pam50 == "LumA"),10,F),
  sample(which(pam50 == "notLumA"),10,F)
	)
holdout <- brca[,rownames(pheno)[idx_holdout]]
colData(holdout)$ID <- as.character(colData(holdout)$patientID)
colData(holdout)$STATUS <- colData(holdout)$pam_mod
tokeep <- setdiff(1:length(pam50),idx_holdout)
brca <- brca[,rownames(pheno)[tokeep]]

pID <- as.character(colData(brca)$patientID)
colData(brca)$ID <- pID
colData(brca)$STATUS <- colData(brca)$pam_mod
```

# Train model on training Samples
The rest of the process for data setup and running the model is identical to that shown in previous vignettes. 

Create feature groupings:
```{r,eval=TRUE}
# genes in mRNA data are grouped by pathways
pathFile <- sprintf("%s/extdata/pathway_ex3.gmt", path.package("netDx"))
pathList <- readPathways(pathFile)

groupList <- list()
groupList[["BRCA_mRNAArray-20160128"]] <- pathList

# clinical data is not grouped; each variable is its own feature
groupList[["clinical"]] <- list(
     age="patient.age_at_initial_pathologic_diagnosis",
	   stage="STAGE"
)
```

Define the function to convert features into patient similarity networks:
```{r,eval=TRUE}
makeNets <- function(dataList, groupList, netDir, ...) {
  netList <- c() # initialize before is.null() check

  # make RNA nets (NOTE: the check for is.null() is important!)
  # (Pearson correlation)
  if (!is.null(groupList[["BRCA_mRNAArray-20160128"]])) {
    netList <- makePSN_NamedMatrix(
        dataList[["BRCA_mRNAArray-20160128"]],
        rownames(dataList[["BRCA_mRNAArray-20160128"]]),
        groupList[["BRCA_mRNAArray-20160128"]],
        netDir, verbose = FALSE,
        writeProfiles = TRUE, runSerially=TRUE, ...)
  }

  # make clinical nets (normalized difference)
  netList2 <- c()
  if (!is.null(groupList[["clinical"]])) {
    netList2 <- makePSN_NamedMatrix(dataList$clinical,
    rownames(dataList$clinical),
    groupList[["clinical"]], netDir,
    simMetric = "custom", 
    customFunc = normDiff, # custom function
    writeProfiles = FALSE,
    sparsify = TRUE, verbose = TRUE, runSerially=TRUE, ...)
  }
  netList <- c(unlist(netList), unlist(netList2))
  return(netList)
}
```

Train the model:

```{r,eval=TRUE}
set.seed(42) # make results reproducible
outDir <- paste(tempdir(),randAlphanumString(),
	"pred_output",sep=getFileSep())
if (file.exists(outDir)) unlink(outDir,recursive=TRUE)
numSplits <- 2L
model <- suppressMessages(
  buildPredictor(
      dataList=brca,groupList=groupList,
      makeNetFunc=makeNets,
      outDir=outDir, 
      numSplits=numSplits,
      featScoreMax=2L,
      featSelCutoff=1L,
      numCores=1L,debugMode=FALSE,
      logging="none")
)
```

As before we get the model results. Let's set a very low threshold for top-scoring features. In practice this would be set higher.
For example, if features are scored out of 10, you could consider a cutoff of features that consistently score 8 out of 10 or higher, for 70% of the trials (`featureSelCutoff=8L`, `featureSelPCt=0.7`)
```{r}
results <- getResults(model,unique(colData(brca)$STATUS),
	featureSelCutoff=1L,featureSelPct=0)
```

Features passing cutoff are here:
```
results$selectedFeatures
```

# Validate on independent samples
Now we use `predict()` to classify samples in the independent dataset. We provide the model with feature design rules in `groupList`, the list of selected features to use in `featSelNet`, the function to convert data into patient similarity networks in `makeNets`, as well as the original and validated datasets in `brca` and `holdout` respectively.

The training data needs to be provided because netDx creates a single patient similarity network with both training and test data. It then uses label propagation to "diffuse" patient labels from training samples to test samples, and labels the latter based on which class they are most similar to.

```{r,eval=TRUE}
outDir <- paste(tempdir(), randAlphanumString(), 
  sep = getFileSep())
if (file.exists(outDir)) unlink(outDir,recursive=TRUE)
dir.create(outDir)

predModel <- suppressMessages(
  predict(brca, holdout, groupList, 
    results$selectedFeatures, makeNets,
    outDir, verbose = FALSE)
)
```

# Plot results of validation

Finally we examine how well our model performed, using `getPerformance()`. 

Compute performance:
```{r,eval=TRUE}
perf <- getPerformance(predModel, c("LumA", "notLumA"))

message(sprintf("AUROC=%1.2f", perf$auroc))
message(sprintf("AUPR=%1.2f", perf$aupr))
message(sprintf("Accuracy=%1.1f%%", perf$acc))
```

We plot the AUROC and AUPR curves using `plotPerf_multi()`.

```{r,eval=TRUE}
plotPerf_multi(list(perf$rocCurve),
  plotTitle = sprintf(
    "BRCA Validation: %i samples", 
    nrow(colData(holdout))))
plotPerf_multi(list(perf$prCurve), 
  plotType = "PR",
  plotTitle = sprintf(
    "BRCA Validation: %i samples", 
    nrow(colData(holdout))))
```

# sessionInfo

```{r,eval=TRUE}
sessionInfo()
```

