---
title: 'Sample NetDx Predictor: Building a predictor using only genetic data'
output:
  html_document: default
  html_notebook: default
---

**Predictor using Kidney renal cell cancer mutation data**

* Yuan et al. 2014 (http://www.nature.com/nbt/journal/v32/n7/full/nbt.2940.html)

**Using patient mutation data, generate similarity networks consisting of patients that share mutations across similar biological pathways. Predict unseen patient subclass based on these similarity networks**

**This example will do the following:**

  * Take patient input data from Yuan et al. and split into train/test groups (80%/20% respectively)
  * Using kidney cancer mutation data for these patients, map mutations to biological pathways for the train set
  * Run feature selection using 10 fold cross validation on these biological pathways to isolate the ones that are most informative
  * Generate patient similarity networks (one for each pathway) and make subclass predictions for the test set based on the feature selected similarity networks

```{r, message=F, warning=F}
# Feature selection for KIRC from PanCancer survival dataset
# 10-fold CV predictor design with mutation data

#Load in the require packages to run the netDx predictor
require(netDx)
require(netDx.examples)


#Set the processing parameters
numCores <- 8L
GMmemory <- 4L
trainProp <- 0.8
cutoff <- 9

#Identify where all the input data is and where to put the output
outRoot <- sprintf("%s/KIRC", getwd())

#Create an output folder with today's date
dt <- format(Sys.Date(),"%y%m%d")
megaDir <- sprintf("%s/KIRC_mut_pathwaysOnly_%s",outRoot,dt)
```

Load in the required packages and set up our input and output
directories
```{r}
#Load in all the pre-formated data required to run the predictor
data(sample_predictor_KIRC_mut_data)
```

* **`data`** will take a set of R variables stored in a 'rda'
file and load them into the current environment

```{r}
#Now we will start to build the classifier

#Create the output directory, delete any directory with the same name
if (file.exists(megaDir)) unlink(megaDir,recursive=TRUE)
dir.create(megaDir)
```


```{r}
#Store all output to the terminal in a log file that we can reference later
logFile <- sprintf("%s/log.txt",megaDir)
sink(logFile,split=TRUE)
#Run the predictor just once
rngNum=1
```
* **rngNum** Is set to 1 here as we only want to run the predictor once using one train/test split. If we wanted to run over several splits, we woulf create a for-loop and iterate over rngNum (ex. for(rngNum in 1:100))

```{r}
	#Create the output directory for this run number
	outDir <- sprintf("%s/rng%i",megaDir,rngNum)
	dir.create(outDir)
	
	#Split patients into train and test groups
	pheno_all$TT_STATUS <-   splitTestTrain(pheno_all,pctT=trainProp,
											  setSeed=rngNum*5)
	
	#Extract just the train set patients to be used for feature selection
	pheno <- subset(pheno_all, TT_STATUS %in% "TRAIN")
	dats_train <- lapply(dats,function(x) {
						 x[,which(colnames(x) %in% pheno$ID)]})
	pat_GR_train <- pat_GR_all[which(pat_GR_all$ID %in% pheno$ID)]
```

* **`splitTestTrain`** will take the input patients and split them into a train and test groups where "trainProp" is the percentage of train patients

```{r}
  #create output directory to store nets
	netDir <- sprintf("%s/networks",outDir)
	pathFile <- sprintf("%s/extdata/Human_160124_AllPathways.gmt",
	   path.package("netDx.examples"))
  #take the pathway information and store it in a data format that is readable by the learner
	pathwayList <- readPathways(pathFile)
	data(genes)
	#store all pathway gene information in genomic ranges objects
	gene_GR     <- GRanges(genes$chrom,IRanges(genes$txStart,genes$txEnd),
	  	name=genes$name2)
	#map genes to pathways which will limit all genes to pathway genes
	path_GRList <- mapNamedRangesToSets(gene_GR,pathwayList)
	names(path_GRList) <- paste("MUT_",names(path_GRList),sep="")

	
	# add somatic mutations at pathway-level
	netList <- makePSN_RangeSets(pat_GR_train, path_GRList, netDir,
		numCores=numCores)
	cat(sprintf("Total of %i nets\n", length(netList)))
	# now create a genemania database
	dbDir	<- GM_createDB(netDir, pheno$ID, outDir,numCores=numCores)


```

* **`makePSN_RangeSets`** Creates patient similarity interaction networks based on range sets

* **`GRanges`** is a genomic rages object, which will store the pathway information in a data structure that allows us to check for overlaps with mutations easily. 

* **`mapNamedRangesToSets`** we have named ranges each
corresponding to genes, and we wish to group these ranges based on metabolic pathway

* **`readPathways`** will take the pathway information and store
it in a data format that is readable by the learner
* The goal here is to set up our pathway information so that the
patient data (mutations in genes) can be mapped to geneomic pathways
```{r}
	# Get the unique subtypes so we can run feature selection once   for each
	subtypes <- unique(pheno$STATUS)
	# run 10-fold cross validation per subtype
	for (g in subtypes) {
    #create a directory for the current subtype
    pDir <- sprintf("%s/%s",outDir,g)
    if (file.exists(pDir)) unlink(pDir,recursive=TRUE)
	  dir.create(pDir)

		cat(sprintf("\n******\nSubtype %s\n",g))
		pheno_subtype <- pheno
		# label patients not in the current class as a "nonpred"
		pheno_subtype$STATUS[which(!pheno_subtype$STATUS %in% g)] <- "nonpred"
		resDir    <- sprintf("%s/GM_results",pDir)
		# query for feature selection comprises of training
		# samples from the class of interest
		trainPred <- pheno_subtype$ID[which(pheno_subtype$STATUS %in% g)]

		# Run Cross validation using the training set of patients that fall under the current subtype
		GM_runCV_featureSet(trainPred, resDir, dbDir$dbDir,
			nrow(pheno_subtype),verbose=T, numCores=numCores,
			GMmemory=GMmemory)

		# patient similarity ranks
		prank <- dir(path=resDir,pattern="PRANK$")
		# network ranks
		nrank <- dir(path=resDir,pattern="NRANK$")
		cat(sprintf("Got %i prank files\n",length(prank)))

	  # Compute network scores
		pTally		<- GM_networkTally(paste(resDir,nrank,sep="/"))
		head(pTally)
		# write network scores to file
		tallyFile	<- sprintf("%s/%s_pathway_CV_score.txt",resDir,g)
		write.table(pTally,file=tallyFile,sep="\t",col=T,row=F,quote=F)
	}
```
The primary goal here is for each phenotype subtype, get the patients that fall under that subtype and run cross validation

* **`GM_runCV_featureSet`** will use GeneMANIA to run cross-validation with a provided subset of networks

* **`GM_networkTally`** will simply tally the score of networks through cross-validation

```{r}
  # Class prediction
	# now create GM databases for each class
	# should contain train + test patients
	# and be limited to nets that pass feature selection
	pheno <- pheno_all
	predRes <- list()
	#Run once for each subtype
	for (g in subtypes) {
		pDir <- sprintf("%s/%s",outDir,g)
		# get feature selected net names
		pTally <- read.delim(
			sprintf("%s/GM_results/%s_pathway_CV_score.txt",pDir,g),
			sep="\t",h=T,as.is=T)
		pTally <- pTally[which(pTally[,2]>=cutoff),1]
		pTally <- sub(".profile","",pTally)
		pTally <- sub("_cont","",pTally)
		cat(sprintf("%s: %i pathways\n",g,length(pTally)))
		netDir <- sprintf("%s/networks",pDir)

    # add somatic mutations at pathway-level
		idx <- which(names(path_GRList) %in% pTally)
		if (any(idx)) {
			cat(sprintf("mutations: included %i nets\n", length(idx)))
			netList3 <- makePSN_RangeSets(pat_GR_all,
				path_GRList[idx],
				netDir,numCores=numCores)
        }

		# create genemania database
		dbDir <- GM_createDB(netDir,pheno$ID,pDir,numCores=numCores)
		# query of all training samples for this class
		qSamps <- pheno$ID[which(pheno$STATUS %in% g &
								 pheno$TT_STATUS%in%"TRAIN")]

		qFile <- sprintf("%s/%s_query",pDir,g)
		#write genemania query files
		GM_writeQueryFile(qSamps,"all",nrow(pheno),qFile)
		#get results file after running genemania queries
		resFile <- runGeneMANIA(dbDir$dbDir,qFile,resDir=pDir)
		#get the ROC curves for the query
		predRes[[g]] <- GM_getQueryROC(sprintf("%s.PRANK",resFile),pheno,g)
	}
```
The goal here is to use the feature selected networks to generate predictions for patient class

* **`GM_writeQueryFile`** will write a GeneMANIA query file based on the input parameters
* **`runGeneMANIA`** runs input queries through GeneMANIA
* **`GM_getQueryROC`** processes the GM PRANK files to get the ROC curve for the query


```{r}
  #Assign patient class by ranking GeneMANIA predictors 
  #and write to file
	predClass <- GM_OneVAll_getClass(predRes)
	out <- merge(x=pheno_all,y=predClass,by="ID")
	outFile <- sprintf("%s/predictionResults.txt",outDir)
	write.table(out,file=outFile,sep="\t",col=T,row=F,quote=F)
  
	#Calculate the accuracy of the predictions by comparing
	#predicted class versus assigned class
	acc <- sum(out$STATUS==out$PRED_CLASS)/nrow(out)
	cat(sprintf("Accuracy on %i blind test subjects = %2.1f%%\n",
		nrow(out), acc*100))
  
	#Generate several performance values and store them for future
	#analysis
	require(ROCR)
	ROCR_pred <- prediction(out$SURVIVEYES_SCORE-out$SURVIVENO,
						out$STATUS=="SURVIVEYES")
	save(predRes,ROCR_pred,file=sprintf("%s/predRes.Rdata",outDir))

```
Our final step will be to assign patient classes to our unseen test set,  calcualte various performance results (accuracy, AUCROC, AUCPR etc) and store both for future analysis

* **`GM_OneVAll_getClass`** will take the GeneMANIA predictors for each class and assign the class with the higher score to each patient
* **`prediction`** will generate several performance values for the prediction results by comparing the assigned class and the predicted class for each patient. We can use these values to determine how well the predictor did.

